Rice yield improvements through plant breeding are offset by inherent yield declines over time

### 2018-03-11 output/review_responses.tex

\textbf{Response:} On the suggestion of several reviewers, we have
moved figure S1 into the main paper.\\

\textbf{Comment: It was important to test whether the interaction cv x
  years after release was significant, because this could give a clue
  as to what is the cause of the yield decline, but it wasn’t
  significant according to the stats which I didn’t really
  understand. Discussion could explore this in more depth; I mean why
  some cvs didn’t lose yield and some did. Maybe only widely adopted
  cvs lost yeild, due to greater presume for disease to evolve
  virulence?}\\

\textbf{Response:} The reviewer should mindful that stating that the
estimated CI for the cultivar-year interaction includes zero is not
the same thing as ``no effect''. Rather, the correct interpretation is
that the effect, if it exists, is smaller than our power to
detect. Therefore, the correct conclusion is that the differences
between cultivars in the decline over time, if existent, is likely
small, though the sign and magnitude of the exact effect is uncertain.

While this is indeed interesting, we have attempted to limit our
speculation on this point given that we do not have sufficient data to
actually test many of these ideas. We have expanded the discussion to
include some of the points raised (lines 237-241). \\

@@

\textbf{Comment: The paper seems to be written to put the blame on the
  cultivars and the challenge on the breeders, the notion of their
  cvs’ yield advantages being “eroded”. This might be the case if a
  biotic stress is evolving virulence against the host plant
  resistance of any cv.  But it is not really the case when there is a
  deterioration in the aspects of the environment (climate, agronomy,
  and/or soil), against which younger cultivars are more resistant or
  tolerant.  For me the 29 kg/ha/yr could be breeding progress in the
  face of the deteriorating E; at about 0.3\% p.a. it is not unlike
  the rate of progress in advanced wheat breeding programs (anything
  near the 1.2\% p.a. which the paper mentions in discussion is very
  rare these days).}\\

\textbf{Response:} The term ``yield erosion'' has been removed due to
these concerns. However, we have taken great pains to specifically
point out that the evidence suggests this inherent to the system (as
the title of this manuscript clearly states), that it is similar to
observations from other rice systems (lines 206,223-228), and that it is
plausibly due to the environmental aspects the reviewer lists (lines
227-236).

@@

However, the evidence from this study explicitly states that given
these data and this analysis, a rate of improvement of 29 kg/ha/yr is
unlikely. The reviewer appears to have missed one of the main
conclusions of the manuscript. We have edited for greater clarity. \\

\textbf{Comment: Really it all depends on the definition of potential
  yield. See van Ittersum and colleagues in FCR Vol 143 March
  2013. Its determined by latest best yielding Cv and the current E in
  the absences of any manageable stresses, both biotic and abiotic,
  and measured in good experiments and models based on such
  experiments. A breakdown in cv resistance is probably best described
  as “erosion”, while a deterioration in the E due, for example, to a
  negative climate trend, is a decrease in the potential yield, but a
  new cv tolerating this deterioration would lift the potential yield
  (see also below).}\\

\textbf{Response:} This is a good point and we have in part addressed
this issue by removing the term “yield erosion”. We have also,
addressed the issue raised by van Ittersum that a changing climate may
change yield potential (lines 233).\\

@@

\textbf{Comment: The authors should read McKay et al (2010, TAG 122,
  225-238) where a similar situation in winter wheat in the UK is
  described. But by having plots with and without fungicide over the
  last 40 years or so, these authors could measure the erosion and
  narrow down the cause (foliar disease), which did vary with
  cultivar.}\\

\textbf{Response:} This is an interesting paper (assuming the reviewer
intended \textbf{Mackay et al. 2011}), but unfortunately these data do
not lend themselves to similar explorations. Unlike Mackay et al., we
do not have paired treated/untreated plots. In the course of writing
this paper, we discussed various means to to narrow down the cause of
yield losses, but ultimately lack the data to do so. We have added a
citation to McKay et al. to the manuscript.\\

\textbf{Comment: Also the authors need to refer to McKenzie et al
  (2014) in the CSSA Special Publicn 33 with a whole section on rice
  yield progress in Ca.}\\

\textbf{Response:} We have addressed this paper more
thoroughly (lines 194-209). Both McKenzie et al. (2014) and Samonte et
al (2014) fail to take yield declines into account in their analyses
and conclusions.  When we exclude these effects, we estimate similar
effects. We show that accounting yield declines results in better
explanation of the data, and resolves appearant contradiction between
short-term and long-term data. \\

@@

\textbf{Comment: These authors show in Fig 10-8 trial results from all
  over the rice areas of Ca, whereas the manuscript deals with trials
  solely from the RES at Briggs.This needs to be discussed and may
  help explain what is happening, as there is little or no evidence in
  McKenzie et al that, for example, the yield of M-202 is declining
  with time.}\\

\textbf{Response:} Explanation added (lines 194-209)\\

\textbf{Comment: What also worries me is that Ca farm yields (and
  Butte county yields) have been increasing since 1995 or so, at a low
  rate (about 70 kg/ha/yr) but increasing, which doesn’t agree with
  Fig S1 no matter what cv you choose.}\\

\textbf{Response:} The reviewer appears to misunderstand basic
differences in the data being presented. We agree and plainly state in
the manuscript that CA farm yields have been increasing (lines
245-255). However, we do not find evidence that gains as large as 70
kg/ha/yr are due to increasing genetic yield potential. It is entirely
possible for farm yields to increase for other reasons which, again,
we outline specifically in the manuscript.\\

@@

\textbf{Comment: I am not sure, however, that you need to show Ca
  yields (Fig 1) because this broadens the discussion a lot, and
  almost demands you look at what is happening across the rice region
  in terms of weather trends etc, probably using a rice simulation
  model (although modelling at just RES itself would be relevant to
  the paper also). It is useful to show how close farm yields are to
  RES trial yields but that could be covered in a text sentence to
  two. And if you want to discuss this bigger picture, please include
  a simple map of rice areas in Ca.}\\

\textbf{Response:} Fig. 1 has been moved to supplemental material.\\

\textbf{Comment: Table 2 lists the key cv data and should be up front
  in the paper. You could add to the Table brief comments on the
  important quality attr ibutes of the various releases (if there any)
  and finally you could include a column showing the important data of
  Fig 2, which has nothing to do with the modelling.}\\

\textbf{Response:} Table 1 and 2 have been swapped. Quality attributes
are not of direct relevance to this paper, so we have provided a
citation for where this information can be found for those interested
(line 261). We disagree that the data from Figure 2 should be
presented in table form. Data of this nature is best presented in
figures, and presenting these data twice is redundent.\\

@@

\textbf{Comment: This is the likely cause: a more thorough exploration
  of biotic stresses is needed : blast, nematodes, etc}\\

\textbf{Response: } As we do not have the data to explore this, this
discussion would be solely speculation. For more, see our response
above.\\

\textbf{Comment: Yes, this would be alarming....a steady decrease in
  the potential of all paddy Es, but why now after millennia of
  paddy?}\\

\textbf{Response: } As we state, more study is needed to verify and
determine the extent and cause of this (lines 241-243).\\

### 2018-03-11 output/breed_paper.Rnw
Final comments.
Adds non-breaking hypens.

\documentclass[12pt,]{elsarticle}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{longtable,booktabs}
\usepackage{comment}
\usepackage{graphicx,grffile, textcomp}
\usepackage{setspace}
\usepackage{etoolbox}
\usepackage{float}
\doublespacing
\usepackage{lineno}
\usepackage{lscape} %% Harvard
\bibliographystyle{model2-names}\biboptions{authoryear}
\pagenumbering{gobble}
%% For figure numbers in supplementary info
%% from http://bytesizebio.net/2013/03/11/adding-supplementary-tables-and-figures-in-latex/
\newcommand{\beginsupplement}{%
  \setcounter{table}{0}
  \renewcommand{\thetable}{S\arabic{table}}%
  \setcounter{figure}{0}
  \renewcommand{\thefigure}{S\arabic{figure}}%
}

\newcommand{\tsups}{\textsuperscript}
\usepackage{Sweave}
\SweaveOpts{prefix.string=graphics/report, png=TRUE}

@@

Certainly one question in looking at this situation is, ``do new
cultivars have increased yield potential''?  Like many breeding
programs, one of the goals of the CCRRF is to increase yields through
breeding cultivars with increased yield potential. However, when the
popular cultivar M-202 was released in 1985, it was reported to yield
an average of 11 Mg ha\tsups{-1}~\citep[14\% moisture content
(MC);][]{m202}. In 2015, the reported average yield of the newest
released cultivar M-209 was 10.8 Mg ha\tsups{-1}~\citep[14\%
MC;][]{m209}. However, in annual breeding reports and cultivar release
announcements, data based on yield performance in side-by-side cultivar
trials show that new cultivars usually out yield the older
cultivars. In other rice systems, a decrease in a cultivar's yield
performance over time has been observed~\citep{Peng1998, Peng2010,
Peng2000, Datta1995}, often termed yield erosion~\citep{Peng1998}. It
has been speculated that this yield erosion is due to the inability of
a cultivar to adapt to changing biotic and abiotic
conditions~\citep{Peng1998, Peng2010, Peng2000, Datta1995}. Yield
erosion may explain the apparent contradiction above, in which
side-by-side yield comparisons show yield improvements in cultivars
while overall there may be little change in yield potential. To
address this issue we tested the hypothesis that if rice cultivar
yields erode over time then apparent yield increases in side-by-side
yield comparison tests will not reflect the true magnitude of yield
changes over time.

@@

There was general agreement between the estimates from the Bayesian
and classical statistical approaches for all models, with the
estimates being similar between similar models. The Bayesian estimates
are presented here due to the inherent flexibility of Bayesian
inference, though the classical estimates can be found in the
supplemental material. Comparing the cultivars' yield during each's
initial time (average 3 years) in the experimental trials just before
being released showed that across cultivars, yields ranged from
\Sexpr{round(plot_data["M207", "mean"], 2)} (M-207) to
\Sexpr{round(plot_data["M205", "mean"], 2)} Mg ha\tsups{-1} (M-205)
(Fig \ref{fig:first_yrs}. Comparing these yields from the time of
release for each cultivar does not provide
evidence of genetic yield improvements during the 33-year study
period. Individual exceptions to this were two cultivars (M-204 and
M-205) that had higher yields than the others.

@@

There was no evidence of consistent differences between cultivars in
the rate of yield erosion (Model 3); although visual evidence suggest
otherwise (Fig. \ref{fig:all_cult}), more data is needed. Based on
this, the models excluding varying rates of erosion by cultivar (i.e.,
Models 1 \& 2) are presented here. The yields of each cultivar over time were
evaluated relative to when the cultivar was released using the full
model (Model 1) which accounts for both yield erosion and yield improvements
over time. The average decrease in yield over time (i.e., yield
erosion) once a cultivar was selected was estimated at
\Sexpr{yrs_coef} kg ha\tsups{-1} year\tsups{-1} (90\% CI:
\Sexpr{RES_CI[2]} to \Sexpr{RES_CI[1]}) (Fig \ref{fig:RES_year}; Table
\ref{tbl:coef}). The estimated 560 kg ha\tsups{-1} yield advantage of
M-206 over M-202 (the current and former predominant cultivars,
respectively), can be attributed solely to the erosion of M-202's
yield over time (Fig. \ref{fig:first_yrs} \&
\ref{fig:compare_yrs}). Further supporting this estimate of yield
erosion, both cultivars yielded similarly at their respective year of
release (Fig. \ref{fig:compare_yrs}), despite the fact that M-202 was
released 19 years prior to M-206 (Table
\ref{tbl:ids}).

@@

In CA, the cultivars with broad adaptation and adoption over the 33
year period of this study are M-202, M-206, and M-105. Cold tolerance
is a major problem for much of the California rice producing region
~\citep{espe2017}, although it is a bigger problem in the southern
portion of the Sacramento Valley due to the influence of the ocean
winds through the Delta ~\citep{espe16a}. These three cultivars have a
higher degree of cold tolerance, high yield potential, and relatively
short duration and are thus broadly grown with over 70\% of the area
in these cultivars at any given time during the study
period~\citep{ucce}. There are several cultivars that were found to
have higher than average yields, even after correcting for yield
erosion. However, these cultivars (M-201, M-204, M-205, and possibly
M-209; Fig. \ref{fig:ids_comp}) are not widely adopted. Importantly,
not all medium grain cultivars that were released were purported to
have higher yields than previous cultivars. Some cultivars had special
traits, but upon release had similar yields to more broadly adapted
cultivars. Examples of this include M-208, which had resistance to
blast disease (\textit{Magnaporthe grisea}) and M-402 which is a
premium quality cultivar. Other cultivars such as M-205 and M-209,
while high yielding, do not have the cold tolerance of the other
cultivars and are thus only recommended for the northern portion of
the Sacramento Valley.

Not accounting for yield erosion, the estimated yield improvement of
\Sexpr{only_release_coef} kg ha\tsups{-1} year\tsups{-1}
(Fig. \ref{fig:wo_erosion}) is lower than the 37 to 60 kg ha\tsups{-1}
year\tsups{-1} estimated by~\citet{Omar2014} in side-by-side
comparisons. However, when accounting for yield erosion and looking
across all cultivars, there is no conclusive evidence of increased
genetic yield potential in medium grain rice cultivars released by
CCRRF's breeding improvement program over the period
\Sexpr{min(mod_data[,'year'])} to \Sexpr{max(mod_data[,'year'])}. 
Grain yield over this period has stayed relatively constant at an average
\Sexpr{round(mean(mod_data[,"yield_kg"])/1000, 1)} Mg
ha\tsups{-1}. Since the estimated yield improvement spans from
\Sexpr{rel_CI[1]} to \Sexpr{rel_CI[2]} kg ha\tsups{-1} year\tsups{-1}
(Fig. \ref{fig:RES_year}; Table \ref{tbl:coef}), both increases and
decreases in the genetic yield potential over the study period are
plausible. This analysis suggests that the breeding program has
maintained but not substantially improved the yield
potential in the face of changing biotic and abiotic pressures. This
conclusion holds true even when evaluating the cultivars which have
been breed for broad adaption and high yield potential (M-202, M-206
and M-105). Yields of these cultivars at time of release have all
yielded between
\Sexpr{tmp=mod_data[mod_data[,"yrs_in_trial"]<=0&mod_data[,"id"]%in%c("M202","M206","M105"),];round(min(by(tmp,tmp[,"id"],function(x)mean(x[,"yield_kg"]))/1000),2)}
and
\Sexpr{round(max(by(tmp,tmp[,"id"],function(x)mean(x[,"yield_kg"]))/1000),2)}
Mg ha\tsups{-1} with no indication of yields increasing
(Fig. \ref{fig:first_yrs}).

@@

Our investigation here is narrowly focused on the rough-rice/field
grain yield. Other advancements due to plant improvement, including
increased quality~\citep{mckenzie1993} and milling
yield~\citep{mckenzie1994}, have been made over the study period and
are important. Additionally, crop duration has been reduced in CA rice
cultivars. For example, while the predominate cultivars M-202 and
M-206 had similar yield potentials at the time of release (Fig. 5),
M-206 matures about 6 days earlier than M-202 (Table
\ref{tbl:ids}). Shorter duration is very important in CA due to an
already short growing season and water limitations (shorter duration
cultivars would presumably require less irrigation water).

California state average yields have increased slowly during the study
period (Fig. \ref{fig:state_yield}) \citep{espe16a}. What are the
potential causes?  It is not due to the adoption of the broadly
adapted cultivars (M-202, M-206, and M-105) as there is no evidence of
increased yield potential among these (Fig. \ref{fig:first_yrs} \&
\ref{fig:compare_yrs}). Although M-205 is not broadly adapted, it has
increase yield potential and its adoption (although limited) in the
northern and warmer part of CA may contribute to some yield gains
reported at the state level. Other factors include improvement in
agronomic management, including more efficient nutrient
management~\citep{Lundy2012, linquist2009}, and weed
management~\citep{Pittelkow2012, linquist2008}. However,
recent rates of state-wide yield gains are far less than those of the
preceding era, consistent with a system near the yield
ceiling~\citep{Grassini2013}.

@@

@
\caption{The performance of two major cultivars over the period
\Sexpr{min(mod_data[,'year'])} -- \Sexpr{max(mod_data[,'year'])}. The
solid line is the estimated yield erosion (\Sexpr{yrs_coef} kg
ha\tsups{-1} year\tsups{-1}), while the dashed line is the published
3-year average yield of M-202 at its release (1982-1984). Both
cultivars have a similar yield at their respective year of release,
and similar decreases over time.}
\label{fig:compare_yrs2}
\end{figure}
\newpage
\end{comment}

>>>>

\documentclass[12pt,]{elsarticle}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{longtable,booktabs}
\usepackage{comment}
\usepackage{graphicx,grffile, textcomp}
\usepackage{setspace}
\usepackage{etoolbox}
\usepackage{float}
\doublespacing
\usepackage{lineno}
\usepackage{lscape} %% Harvard
\usepackage[shortcuts]{extdash} %% Avoid breaking on M--20X
\bibliographystyle{model2-names}\biboptions{authoryear}
\pagenumbering{gobble}
%% For figure numbers in supplementary info
%% from http://bytesizebio.net/2013/03/11/adding-supplementary-tables-and-figures-in-latex/
\newcommand{\beginsupplement}{%
  \setcounter{table}{0}
  \renewcommand{\thetable}{S\arabic{table}}%
  \setcounter{figure}{0}
  \renewcommand{\thefigure}{S\arabic{figure}}%
}

\newcommand{\tsups}{\textsuperscript}
\usepackage{Sweave}
\SweaveOpts{prefix.string=graphics/report, png=FALSE}

@@

Certainly one question in looking at this situation is, ``do new
cultivars have increased yield potential''?  Like many breeding
programs, one of the goals of the CCRRF is to increase yields through
breeding cultivars with increased yield potential. However, when the
popular cultivar M\=/202 was released in 1985, it was reported to yield
an average of 11 Mg ha\tsups{-1}~\citep[14\% moisture content
(MC);][]{m202}. In 2015, the reported average yield of the newest
released cultivar M\=/209 was 10.8 Mg ha\tsups{-1}~\citep[14\%
MC;][]{m209}. However, in annual breeding reports and cultivar release
announcements, data based on yield performance in side-by-side cultivar
trials show that new cultivars usually out yield the older
cultivars. In other rice systems, a decrease in a cultivar's yield
performance over time has been observed~\citep{Peng1998, Peng2010,
Peng2000, Datta1995}, often termed yield erosion~\citep{Peng1998}. It
has been speculated that this yield erosion is due to the inability of
a cultivar to adapt to changing biotic and abiotic
conditions~\citep{Peng1998, Peng2010, Peng2000, Datta1995}. Yield
erosion may explain the apparent contradiction above, in which
side-by-side yield comparisons show yield improvements in cultivars
while overall there may be little change in yield potential. To
address this issue we tested the hypothesis that if rice cultivar
yields erode over time then apparent yield increases in side-by-side
yield comparison tests will not reflect the true magnitude of yield
changes over time.

@@

There was general agreement between the estimates from the Bayesian
and classical statistical approaches for all models, with the
estimates being similar between similar models. The Bayesian estimates
are presented here due to the inherent flexibility of Bayesian
inference, though the classical estimates can be found in the
supplemental material. Comparing the cultivars' yield during each's
initial time (average 3 years) in the experimental trials just before
being released showed that across cultivars, yields ranged from
\Sexpr{round(plot_data["M207", "mean"], 2)} (M\=/207) to
\Sexpr{round(plot_data["M205", "mean"], 2)} Mg ha\tsups{-1} (M\=/205)
(Fig \ref{fig:first_yrs}. Comparing these yields from the time of
release for each cultivar does not provide
evidence of genetic yield improvements during the 33-year study
period. Individual exceptions to this were two cultivars (M\=/204 and
M\=/205) that had higher yields than the others.

@@

There was no evidence of consistent differences between cultivars in
the rate of yield erosion (Model 3); although visual evidence suggest
otherwise (Fig. \ref{fig:all_cult}), more data is needed. Based on
this, the models excluding varying rates of erosion by cultivar (i.e.,
Models 1 \& 2) are presented here. The yields of each cultivar over time were
evaluated relative to when the cultivar was released using the full
model (Model 1) which accounts for both yield erosion and yield improvements
over time. The average decrease in yield over time (i.e., yield
erosion) once a cultivar was selected was estimated at
\Sexpr{yrs_coef} kg ha\tsups{-1} year\tsups{-1} (90\% CI:
\Sexpr{RES_CI[2]} to \Sexpr{RES_CI[1]}) (Fig \ref{fig:RES_year}; Table
\ref{tbl:coef}). The estimated 560 kg ha\tsups{-1} yield advantage of
M\=/206 over M\=/202 (the current and former predominant cultivars,
respectively), can be attributed solely to the erosion of M\=/202's
yield over time (Fig. \ref{fig:first_yrs} \&
\ref{fig:compare_yrs}). Further supporting this estimate of yield
erosion, both cultivars yielded similarly at their respective year of
release (Fig. \ref{fig:compare_yrs}), despite the fact that M\=/202 was
released 19 years prior to M\=/206 (Table
\ref{tbl:ids}).

@@

In CA, the cultivars with broad adaptation and adoption over the 33
year period of this study are M\=/202, M\=/206, and M\=/105. Cold tolerance
is a major problem for much of the California rice producing region
~\citep{espe2017}, although it is a bigger problem in the southern
portion of the Sacramento Valley due to the influence of the ocean
winds through the Delta ~\citep{espe16a}. These three cultivars have a
higher degree of cold tolerance, high yield potential, and relatively
short duration and are thus broadly grown with over 70\% of the area
in these cultivars at any given time during the study
period~\citep{ucce}. There are several cultivars that were found to
have higher than average yields, even after correcting for yield
erosion. However, these cultivars (M\=/201, M\=/204, M\=/205, and possibly
M\=/209; Fig. \ref{fig:ids_comp}) are not widely adopted. Importantly,
not all medium grain cultivars that were released were purported to
have higher yields than previous cultivars. Some cultivars had special
traits, but upon release had similar yields to more broadly adapted
cultivars. Examples of this include M\=/208, which had resistance to
blast disease (\textit{Magnaporthe grisea}) and M\=/402 which is a
premium quality cultivar. Other cultivars such as M\=/205 and M\=/209,
while high yielding, do not have the cold tolerance of the other
cultivars and are thus only recommended for the northern portion of
the Sacramento Valley.

Not accounting for yield erosion, the estimated yield improvement of
\Sexpr{only_release_coef} kg ha\tsups{-1} year\tsups{-1}
(Fig. \ref{fig:wo_erosion}) is lower than the 37 to 60 kg ha\tsups{-1}
year\tsups{-1} estimated by~\citet{Omar2014} in side-by-side
comparisons. However, when accounting for yield erosion and looking
across all cultivars, there is no conclusive evidence of increased
genetic yield potential in medium grain rice cultivars released by
CCRRF's breeding improvement program over the period
\Sexpr{min(mod_data[,'year'])} to \Sexpr{max(mod_data[,'year'])}. 
Grain yield over this period has stayed relatively constant at an average
\Sexpr{round(mean(mod_data[,"yield_kg"])/1000, 1)} Mg
ha\tsups{-1}. Since the estimated yield improvement spans from
\Sexpr{rel_CI[1]} to \Sexpr{rel_CI[2]} kg ha\tsups{-1} year\tsups{-1}
(Fig. \ref{fig:RES_year}; Table \ref{tbl:coef}), both increases and
decreases in the genetic yield potential over the study period are
plausible. This analysis suggests that the breeding program has
maintained but not substantially improved the yield
potential in the face of changing biotic and abiotic pressures. This
conclusion holds true even when evaluating the cultivars which have
been breed for broad adaption and high yield potential (M\=/202, M\=/206
and M\=/105). Yields of these cultivars at time of release have all
yielded between
\Sexpr{tmp=mod_data[mod_data[,"yrs_in_trial"]<=0&mod_data[,"id"]%in%c("M202","M206","M105"),];round(min(by(tmp,tmp[,"id"],function(x)mean(x[,"yield_kg"]))/1000),2)}
and
\Sexpr{round(max(by(tmp,tmp[,"id"],function(x)mean(x[,"yield_kg"]))/1000),2)}
Mg ha\tsups{-1} with no indication of yields increasing
(Fig. \ref{fig:first_yrs}).

@@

Our investigation here is narrowly focused on the rough-rice/field
grain yield. Other advancements due to plant improvement, including
increased quality~\citep{mckenzie1993} and milling
yield~\citep{mckenzie1994}, have been made over the study period and
are important. Additionally, crop duration has been reduced in CA rice
cultivars. For example, while the predominate cultivars M\=/202 and
M\=/206 had similar yield potentials at the time of release (Fig. 5),
M\=/206 matures about 6 days earlier than M\=/202 (Table
\ref{tbl:ids}). Shorter duration is very important in CA due to an
already short growing season and water limitations (shorter duration
cultivars would presumably require less irrigation water).

California state average yields have increased slowly during the study
period (Fig. \ref{fig:state_yield}) \citep{espe16a}. What are the
potential causes?  It is not due to the adoption of the broadly
adapted cultivars (M\=/202, M\=/206, and M\=/105) as there is no evidence of
increased yield potential among these (Fig. \ref{fig:first_yrs} \&
\ref{fig:compare_yrs}). Although M\=/205 is not broadly adapted, it has
increase yield potential and its adoption (although limited) in the
northern and warmer part of CA may contribute to some yield gains
reported at the state level. Other factors include improvement in
agronomic management, including more efficient nutrient
management~\citep{Lundy2012, linquist2009}, and weed
management~\citep{Pittelkow2012, linquist2008}. However,
recent rates of state-wide yield gains are far less than those of the
preceding era, consistent with a system near the yield
ceiling~\citep{Grassini2013}.

@@

@
\caption{The performance of two major cultivars over the period
\Sexpr{min(mod_data[,'year'])} -- \Sexpr{max(mod_data[,'year'])}. The
solid line is the estimated yield erosion (\Sexpr{yrs_coef} kg
ha\tsups{-1} year\tsups{-1}), while the dashed line is the published
3-year average yield of M\=/202 at its release (1982-1984). Both
cultivars have a similar yield at their respective year of release,
and similar decreases over time.}
\label{fig:compare_yrs2}
\end{figure}
\newpage
\end{comment}
